---
title: "Massachusetts Bay Transportation Authority Ridership Data Cleaning"
author: "Surabhi Chouhan"
date: "January 11, 2017"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Import packages
```{r}
library("readxl")
library("tidyr")
library("ggplot2")
```


```{r, echo=FALSE}
mbta <- read_excel("D:/Surabhi docs/portfolio/data cleaning/mbta.xlsx", skip=1)
str(mbta)
head(mbta)
summary(mbta)
```

##Now time for some data cleaning
```{r, echo=FALSE}
mbta2 <- mbta[c(-1,-7,-11),]

mbta3 <- mbta2[,-1]
```

##In this data, most typical problem of messy data can be seen, that is, "variables are stored in rows instead of columns", so firstly to correct that-
```{r, echo=FALSE}
mbta4 <- gather(mbta3, month, thousand_riders, -mode)
head(mbta4)
```

##The column "thousand_riders" showing the average weekday number of riders, is a character vector so first step will be to coerce it into numeric values to make further calculations easy
```{r, echo=FALSE}
mbta4$thousand_riders <- as.numeric(mbta4$thousand_riders)
head(mbta4)
```

##Second basic symptomof messy data is, variables stored as rows, so to make this right we have to spread the variables into columns
```{r, echo=FALSE}
mbta5 <- spread(mbta4, mode, thousand_riders)
head(mbta5)
```

##Divide the month column containing month-year data together into "Month and "Year" column separately
```{r, echo=FALSE}
mbta6 <- separate(mbta5, month, c("Year", "Month"), sep="-")
head(mbta6)
```

##Before the data as "clean", let's look for any obvious mistakes or outliers in the data
```{r, echo=FALSE}
head(mbta6)
```


##By looking at the data entries in the column "Boat", we can see that there seems to be an obvious data entry error here - Easily seen with a histogram
```{r, echo=FALSE}
hist(mbta6$Boat)
```

##Now to deal with this entry error-
##First find out the row in which this erroneus entry is present and replace the "40" with a "4"
```{r, echo=FALSE}
i <- which(mbta6$Boat == 40)
mbta6$Boat[i] <- 4
```

##Now generate a histogram of Boat column
```{r, echo=FALSE}
hist(mbta6$Boat)
```

##Now that our data is clean lets compare the messy data we got and the clean data, there is quite a lot of visible difference
```{r, echo=FALSE}
head(mbta)
head(mbta6)
```


##Now lets do some visualizations to finally present the clean data
##Look at Boat and Trackless Trolley ridership over time
```{r, echo=FALSE}
mbta_boat <- subset(mbta4, mbta4$mode == "Boat" | mbta4$mode == "Trackless Trolley")

ggplot(mbta_boat, aes(x = month, y = thousand_riders, col = mode)) +  
  geom_point() + 
  scale_x_discrete(name = "Month", breaks = c("2007-01", "2008-01", "2009-01", "2010-01", "2011-01")) +        scale_y_continuous(name = "Avg Weekday Ridership (thousands)")
```

##Look at all T ridership over time
```{r, echo=FALSE}
ggplot(mbta4, aes(x = month, y = thousand_riders, col = mode)) + geom_point() + 
  scale_x_discrete(name = "Month", breaks = c("2007-01", "2008-01", "2009-01", "2010-01", "2011-01")) +  
  scale_y_continuous(name = "Avg Weekday Ridership (thousands)")
```


