---
title: "US School Attendence Data Cleaning"
author: "Surabhi Chouhan"
date: "January 12, 2017"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r, warning=FALSE}
library("gdata")
library("stringr")
library("dplyr")
```


#import data and exploring and understanding the data
```{r}
att <- read.xls("D:/Surabhi docs/portfolio/data cleaning/attendance.xls", perl= "C:/Strawberry/perl/bin/perl.exe")

colnames(att)
head(att)
tail(att)
str(att)
```

##Get rid of all the unnecessary rows
```{r}
rem_row <- c(3,56,57,58,59)
att2 <- att[-(rem_row),]
```

##Get rid of unnecessary columns
```{r}
#Does not contain attendence data
rem_col <-  c(3,5,7,9,11,13,15,17)
att3 <- att2[,-(rem_col)]
```

##Splitting the data - to separate the data of elementary schools, secondary schools and all schools
```{r}
# Subset just elementary schools: att_elem
att_elem <- att3[,c(1,6,7)]

# Subset just secondary schools: att_sec
att_sec <- att3[,c(1,8,9)]

# Subset all schools: att4
att4 <- att3[,c(1:5)]
```

##Replacing the names of columns
```{r}
cnames <- c("state", "avg_attend_pct", "avg_hr_per_day", 
            "avg_day_per_yr", "avg_hr_per_yr")

colnames(att4) <- cnames

# Remove first two rows of att4: att5 as it does not contain any useful data
att5 <- att4[-(1:2),]
colnames(att5)
```

##One of the characteristic of this messy data is that in gthe state names, extra characters have been added to make the length of all the state names to be same. We need to remove those extra characters to make the data useful for us
```{r}
# Remove all periods in state column
att5$state <- str_replace_all(att5$state, "\\.", "")

# Remove white space around state names    
att5$state <- str_trim(att5$state)

head(att5)
```

##Some final type conversions
```{r}
# Define vector containing numerical columns: cols
cols <- c(2:5)

# Use sapply to coerce cols to numeric
att5[, cols] <- sapply(att5[, cols], as.numeric)
```

##Now that our data is clean, lets take a look at how the data looked like originally and how it looks after cleaning
```{r}
#original messy data
str(att)

#clean data
str(att5)
```



















